\section{Summary}
In this work, we have developed a comprehensive study of quantum thermometry, focusing on both equilibrium and dynamical regimes. After building the theoretical foundation and optimal bounds, we evaluated two physically motivated models: the Star and the All-to-All configurations. These models appeared in recent works after optimizing all possible two-body spin network configurations. The analysis showed that the Star Model approaches the optimal bound at large system sizes, while the All-to-All Model performs better for small probes, despite its longer thermalization times.

We then extended our analysis to dynamical quantum thermometry, to develop new results in the field by simulating thermalization via Lindblad master equations. We observed that energy reaches equilibrium relatively quickly, while the QFI — being more sensitive to fine-grained features of the population distribution — converges more slowly. This highlights a trade-off between thermometric sensitivity and thermalization time that becomes more pronounced with increasing system size.

Overall, this work combines analytical insights with numerical simulations to clarify how structure and dynamics affect the performance of quantum thermometers. These findings highlight practical limitations in out-of-equilibrium thermometry and can inform the design of efficient quantum sensors for applications in low-temperature physics, quantum heat engines, and non-equilibrium thermodynamics.

\section{Outlook (or future and actual work)}
Recent developments in quantum thermometry suggest several promising directions for extending this work. One is exploring optimal finite-time estimation strategies using local controls or coarse-grained measurements, which may enhance practicality without sacrificing precision. Incorporating adaptive schemes or periodically driven dynamics could also improve sensitivity in the transient regime. On the experimental side, implementing spin-network thermometers in platforms like trapped ions or superconducting qubits is becoming increasingly feasible. Finally, integrating thermometric tasks with quantum heat engines — a direction explored by Perarnau-Llobet and collaborators — offers a path toward multifunctional quantum devices that combine sensing and thermodynamic processing.
